---
title: Cnidarian hair cell development illuminates an ancient role for the class IV
  POU transcription factor in defining mechanoreceptor identity
author: "Pablo Yamild Rosiles Loeza"
date: "9/27/2021"
output: html_document
---


### NOTES: 

+ Within the code, there are square brackets in square brackets those paths that must be modified by the user for its use.

+ The lines of code marked with a @ between the lines must be run repeatedly for each sample. 

# DOWNLOAD DATA

### Reference genome data 
To initiate the analysis, we required to process the RNA-seq data from Tournière, et al. (2020).
To achieve this, reference genomic data must be downloaded.

To download the complete fasta file of the sequence genome. Run in terminal
```{bash, eval=F}
wget ftp://ftp.ensemblgenomes.org/pub/metazoa/release-51/fasta/nematostella_vectensis/dna/Nematostella_vectensis.ASM20922v1.dna.toplevel.fa.gz
```

To download all annotation files, please go to https://doi.org/https://doi.org/10.6084/m9.figshare.807696.v2 
Or run in terminal

```{bash, eval=F}
wget https://figshare.com/ndownloader/articles/807696/versions/2
unzip 2
```

In order to run this script, you can clone the repository from Github. The annotation data downloaded in the previous step should be located in the folder ref_genomes


### RNA-seq data 

To download all files interactively, go to https://www.ebi.ac.uk/arrayexpress/experiments/E-MTAB-8658/ Or run in terminal
```{bash, eval=F}
cd data
# POU mutants data
wget ftp://ftp.sra.ebi.ac.uk/vol1/fastq/ERR380/003/ERR3809533/ERR3809533.fastq.gz
wget ftp://ftp.sra.ebi.ac.uk/vol1/fastq/ERR380/002/ERR3809532/ERR3809532.fastq.gz
wget ftp://ftp.sra.ebi.ac.uk/vol1/fastq/ERR380/004/ERR3809534/ERR3809534.fastq.gz
wget ftp://ftp.sra.ebi.ac.uk/vol1/fastq/ERR380/005/ERR3809535/ERR3809535.fastq.gz

# POU wild type data
wget ftp://ftp.sra.ebi.ac.uk/vol1/fastq/ERR380/006/ERR3809536/ERR3809536.fastq.gz
wget ftp://ftp.sra.ebi.ac.uk/vol1/fastq/ERR380/007/ERR3809537/ERR3809537.fastq.gz
wget ftp://ftp.sra.ebi.ac.uk/vol1/fastq/ERR380/008/ERR3809538/ERR3809538.fastq.gz
wget ftp://ftp.sra.ebi.ac.uk/vol1/fastq/ERR380/009/ERR3809539/ERR3809539.fastq.gz

```


To uncompress the file, run 
```{bash,eval=F}
gzip *.fastq.gz
#Then we move the files
mv *.fastq /before_trim
```

These data downloaded should be located in data/before_trim folder

### ChIP-seq data

Go to [link] Or run in terminal
```{bash,eval=F}
 wget [link]
```


# DATA QUALITY

### RNA-seq data

For RNA-seq data, we use fastqc, in terminal run
```{bash, eval=F}
fastqc -O quality_files/before_trim data/*fastq

```

The html files, will be located in quality_files/before_trim

Since k-mer content, sequence duplication levels and overrepresented sequences may fail due to adaptor content or poor sequence quality, we used the Trimmomatic tool to process the datasets.
For this, we download illumina adapters from https://gist.github.com/photocyte/3edd9401d0b13476e60f8b104c2575f8 and then run in terminal

```{bash,eval=F}

cd data/
java -jar [#PATH_TO_TRIMMOMATIC_JAVA_FILE]/trimmomatic-0.32.jar SE -threads 4 before_trim/ERR3809539.fastq Nematostella_vectensis_CH2_12d_POU-+_wild_rep4.fastq ILLUMINACLIP:Sequencing_adaptors.fasta:2:30:10 SLIDINGWINDOW:4:15 MINLEN:60 HEADCROP:10
java -jar [#PATH_TO_TRIMMOMATIC_JAVA_FILE]/trimmomatic-0.32.jar SE -threads 4 before_trim/ERR3809538.fastq Nematostella_vectensis_CH2_12d_POU-+_wild_rep3.fastq ILLUMINACLIP:Sequencing_adaptors.fasta:2:30:10 SLIDINGWINDOW:4:15 MINLEN:60 HEADCROP:10
java -jar [#PATH_TO_TRIMMOMATIC_JAVA_FILE]/trimmomatic-0.32.jar SE -threads 4 before_trim/ERR3809537.fastq Nematostella_vectensis_CH2_12d_POU-+_wild_rep2.fastq ILLUMINACLIP:Sequencing_adaptors.fasta:2:30:10 SLIDINGWINDOW:4:15 MINLEN:60 HEADCROP:10
java -jar [#PATH_TO_TRIMMOMATIC_JAVA_FILE]/trimmomatic-0.32.jar SE -threads 4 before_trim/ERR3809536.fastq Nematostella_vectensis_CH2_12d_POU-+_wild_rep1.fastq ILLUMINACLIP:Sequencing_adaptors.fasta:2:30:10 SLIDINGWINDOW:4:15 MINLEN:60 HEADCROP:10

java -jar [#PATH_TO_TRIMMOMATIC_JAVA_FILE]/trimmomatic-0.32.jar SE -threads 4 before_trim/ERR3809533.fastq Nematostella_vectensis_CH2_12d_POU--_rep1 ILLUMINACLIP:Sequencing_adaptors.fasta:2:30:10 SLIDINGWINDOW:4:15 MINLEN:60 HEADCROP:10
java -jar [#PATH_TO_TRIMMOMATIC_JAVA_FILE]/trimmomatic-0.32.jar SE -threads 4 before_trim/ERR3809532.fastq Nematostella_vectensis_CH2_12d_POU--_rep2 ILLUMINACLIP:Sequencing_adaptors.fasta:2:30:10 SLIDINGWINDOW:4:15 MINLEN:60 HEADCROP:10
java -jar [#PATH_TO_TRIMMOMATIC_JAVA_FILE]/trimmomatic-0.32.jar SE -threads 4 before_trim/ERR3809534.fastq Nematostella_vectensis_CH2_12d_POU--_rep3 ILLUMINACLIP:Sequencing_adaptors.fasta:2:30:10 SLIDINGWINDOW:4:15 MINLEN:60 HEADCROP:10
java -jar [#PATH_TO_TRIMMOMATIC_JAVA_FILE]/trimmomatic-0.32.jar SE -threads 4 before_trim/ERR3809535.fastq Nematostella_vectensis_CH2_12d_POU--_rep4 ILLUMINACLIP:Sequencing_adaptors.fasta:2:30:10 SLIDINGWINDOW:4:15 MINLEN:60 HEADCROP:10
```

We included the code for all replicates in order to ensure that the file names are correct.We can check data quelity again, using this code in terminal

```{bash,eval=F}
fastqc -O quality_files/after_trim data/*fastq

```

The html files, will be located in quality_files/after_trim


### ChIP-seq data
We use ChipQC in R

You can install it, using 
```{r,eval=FALSE}
BiocManager::install("ChIPQC")
```

And then, we can run 
```{r}
library(ChIPQC)
sample1 = ChIPQCsample('../data/[]')
sample2 = ChIPQCsample('../data/[]')
sample3 = ChIPQCsample('../data/[]')

# Report from sample 1
ChIPQCreport(sample1)
# Report from sample 2
ChIPQCreport(sample2)
# Report from sample 3
ChIPQCreport(sample3)
```



# GUIDED SEQUENCE ALIGNMENT 

We proceeded to the alignment of the reads with the reference genome using STAR. 
First, we need the headers to be equal, to achieve this, run in data folder
```{bash, eval=F}
sed 's/NEMVE//g' [fasta_file_with_sequences] > Nvectensis.fa
```

Then we generate the index using the following lines in terminal 
```{bash, eval=F}
STAR --runThreadN 6 --runMode genomeGenerate --genomeDir STAR_genome_index --genomeFastaFiles ref_genomes/[Nvectensis.fa] --sjdbGTFfile ref_genomes/[nveGenes.vienna130208.nemVec1.gtf] --sjdbOverhang 99
```

Once the index was completed, we used each trimmed read and aligned to the reference genome using STAR tools. The following code is an example of the code for replicate 1 of the POU+/- biological sample. 

```{bash,eval=F}
#@ run for each replicate
STAR --runThreadN 6 --genomeDir  ref_genomes/ --readFilesIn data/Nematostella_vectensis_CH2_12d_POU-+_wild_rep1.fastq --outFileNamePrefix alignment/Nematostella_vectensis_CH2_12d_POU-+_wild_rep1
#@ run for each replicate
```

# RNA-seq

First we must sort and count the alignments made by STAR. For this, we use HTseq and samtools. In terminal, we run 

```{bash, eval=FALSE}
#@ run for each replicate
# To compress file to bam
samtools view -S -b alignment/Nematostella_vectensis_CH2_12d_POU-+_wild_rep1.out.sam > Nvect_POU-+_rep1_Aligned.bam
# Sort 
samtools sort Nvect_POU-+_rep1_Aligned.bam -o Nvect_POU-+_rep1_sorted.bam
# Counts 
htseq-count -f bam -s yes -r pos Nvect_POU-+_rep1_sorted.bam ../ref_genomes/[nveGenes.vienna130208.nemVec1.gtf] > POU-+_rep1.counts
#@ run for each replicate
```

Finally, merge the readings into a single count table. 



# DATA QUALITY

# SYSTEM INFORMATION


The system information used to run this analysis.
```{r}
options(width = 120)
pkgs <- loadedNamespaces()
pkgs <- installed.packages()[, "Package"]
sessioninfo::session_info(pkgs)
```




